State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-05-18 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
edits to freq analysis and state estimation

	\label{DiscreteTimeModel} v_{t+T_s}\left(\mathbf{r}\right) &=& \xi v_t\left(\mathbf{r}\right) + T_s \int_\Omega { w\left(\mathbf{r}-\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'} \nonumber\\ 

@@

	 w\left(\mathbf{r}-\mathbf{r}'\right) =\boldsymbol{\psi}^\top\left(\mathbf{r}-\mathbf{r}'\right) \boldsymbol{\theta},

@@

where $\boldsymbol{\psi}(r-r')$ is a vector of Gaussian basis functions and $\boldsymbol{\theta}$ is a vector of scaling parameters. We will assume that we know the parametric form of the connectivity basis functions, where the scaling parameters $\boldsymbol{\theta}$ are unknown. Each connectivity basis function can, individually, be considered a layer in the Wilson and Cowan model, representing short range excitation, surround inhibition and mid-range excitation. Making substitutions of equations~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} into~\ref{DiscreteTimeModel} we get 

@@

Let the spectral representation of the dynamic field at time $t$ be denoted by $V_t(\boldsymbol{\nu})$. Based on Shannon's sampling theorem, for an accurate representation of this field by spatially discrete observations, $V_t(\boldsymbol{\nu})$ needs to be spatially band-limited. Such a condition ensures that $V_t(\boldsymbol{\nu}) \approx 0 ~ \forall \boldsymbol{\nu} > \boldsymbol{\nu}_c$, where $\boldsymbol{\nu}_c$ is a cutoff frequency with  $\boldsymbol{\nu}_c = [\nu_c ~ \nu_c]^\top$. Given such a band-limited field, the distance between adjacent sensors, $\Delta_s$, needs to satisfy 
\begin{equation}
	\label{eq:MinimumSensorDistance} \Delta_s < \frac{1}{2\rho_s\nu_{c}}, 
\end{equation}
where $\rho_s \in \mathbb{R} \ge 1$ is an oversampling parameter. This condition must be satisfied to avoid spectral aliasing effects in the representation of the hidden dynamic field $v_t(\mathbf{r})$ by the sampled observations, $\mathbf{y}_t$.

@@

% 
% Due to the high dimensionality of the brain and the current electrode systems, we can not expect to have more sensors than basis functions.
\section{State and Parameter Estimation}\label{StateAndParameterEstimationSection} In this section we describe the procedure for estimating the states, $\mathbf{x}_t$, and the connectivity kernel parameters, $\boldsymbol \theta$, and the synaptic time constant, $\xi$. The estimation process is a two part iterative algorithm, consisting of a state estimation step followed by a parameter estimation step. At each iteration, the sequence of estimated states is used to update the parameter set. Then the resulting parameters are used to estimate a new state vector for the next iteration. The procedure stops when the parameters converge. The algorithm is initialised using a bounded random state sequence that guarantees the initial estimated parameter set forms a stable kernel.

The unscented Rauch Tung Striebel Smoother (URTSS)~\cite{Sarkka2010} is used for the state estimation. The URTSS incorporates an unscented Kalman Filter (UKF)~\cite{Julier1997, Merwe2003} in a forward iteration to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward iteration to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and the second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. The sigma points, $\mathcal X_i$, are calculated using the unscented transform as follows:
\begin{equation}\label{eq:sigmapoints1}
	\mathcal X_{0}=\bar x 
\end{equation}
\begin{equation}
	\mathcal X_{i}=\bar x+(\sqrt{( L + \lambda)\mathbf P_x})_i, \quad i=1, \dots, L 
\end{equation}
\begin{equation}\label{eq:sigmapoints2}
	\mathcal X_{i}=\bar x-(\sqrt{( L + \lambda)\mathbf P_x})_{i- L}, \quad i= L+1, \dots, 2 L 
\end{equation}
where $\bar x$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$, $\mathbf{P}_x$ is the corresponding covariance matrix from the filtering or smoothing, $(\sqrt{( L + \lambda)\mathbf P_x})_i$ is the $i$th column of the weighted matrix square root of $\mathbf P_x$ and $L$ is the dimension of the state space. The total number of sigma points is $2L+1$. The scaling parameter $\lambda$ is defined as 
\begin{equation}\label{eq:sigmapoints3}
	\lambda=\alpha^2( L+\kappa) - L 
\end{equation}
where the constants $\alpha$ determines the spread of the sigma points, $\beta$ incorporates prior knowledge of the distribution of $\mathbf{x}$ and $\kappa$ is a scaling parameter (see ~\cite{Haykin2001} for more details).

The sigma vectors are propagated through the system equations and weighted to form the predicted mean and covariance. The weights are calculated by 
\begin{equation}
	\mathbf W_0^{(m)}=\frac{\lambda}{ L+\lambda} 
\end{equation}
\begin{equation}
	\mathbf W_0^{(c)}=\frac{\lambda}{ L+\lambda}+(1-\alpha^2+\beta) 
\end{equation}
\begin{equation}
	\mathbf W_i^{(m)}=\mathbf W_i^{(c)}=\frac{1}{2( L+\lambda)} \quad i=1, \dots, 2L. 
\end{equation}
Since the observation equation is linear (equation~\ref{ObservationEquation}) the standard Kalman Filter update equations are  used to correct the predicted states. The state estimates from the forward filtering are used to form a new set of sigma points for the smoother, as described above. To compute the smoother gain, the predicted cross-covariance matrix of the states is also calculated. A summary of the URTSS procedure is given in Algorithm~\ref{UKFAlgorithm}. It should be noted that as the disturbance, $e_t\left(\mathbf{r}\right)$ and the measurement noise, $\boldsymbol{\varepsilon}_t$, are additive. Therefore, the additive form of the URTS is used, otherwise the state vector should be augmented with the noise terms. The additive form reduces the dimension and the number of sigma points used in the algorithm.

\begin{algorithm}
	\begin{small}
	\caption{The Unscented RTS Smoother}\label{UKFAlgorithm} 
	\begin{algorithmic}[1] 
		\State Forward initialisation 
		\begin{equation*}
		 \hat{\mathbf x}_0, \mathbf P_0 
		\end{equation*}
		\State Forward iteration: for $t \in \left\lbrace 0,\cdots, T\right\rbrace $,
		calculate the sigma points $\mathcal X_{i,t}^f$ using equations \ref{eq:sigmapoints1}-\ref{eq:sigmapoints3} and propagate through equation~\ref{eq:QmatrixForSigmapoints}
% 		\begin{small}
		\begin{equation*}
			\mathcal X_{i,t+1}^{f-}=Q(\mathcal X_{i,t}^f) 
		\end{equation*}
% 		\end{small}
		calculate the predicted state and the predicted covariance matrix
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{f-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{f-} 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t +1}^{f-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})^\top+\boldsymbol \Sigma_e 
		\end{equation*}
		compute the filter gain, the filtered state and the filtered covariance matrix using the standard Kalman Filter update equations
		\begin{equation*}
			\mathcal K_{t+1}=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_v)^{-1} 
		\end{equation*}
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{f}=\hat{\mathbf x}_{t+1}^{f-}+\mathcal K_{t+1}(\mathbf y_{t+1}-\mathbf C\hat{\mathbf x}_{t +1}^{f-}) 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t+1}^f=(\mathbf I - \mathcal K_{t+1}\mathbf C)\mathbf P_{t +1}^{f-} 
		\end{equation*}
		\State Backward initialisation 
		\begin{equation*}
			\mathbf P_T^b= \mathbf P_T^f, \quad \hat{\mathbf x}^b_T= \hat{\mathbf x}^f_T 
		\end{equation*}
		\State Backward iteration: for $t \in \left\lbrace T-1, \cdots, 0 \right\rbrace $ calculate the sigma points $\mathcal X_{i,t}^b$ and propogate them through equation \ref{eq:QmatrixForSigmapoints}
		\begin{equation*}
			\mathcal X_{i,t+1}^{b-}=Q(\mathcal X_{i,t}^b,u_t) 
		\end{equation*}
		 compute the predicted state, the predicted covariance matrix and the cross-covariance matrix
		\begin{equation*}
			\hat{\mathbf x}_{t+1}^{b-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{b-} 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t +1}^{b-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})^\top+\boldsymbol \Sigma_e 
		\end{equation*}
		\begin{equation*}
			\mathbf M_{t +1}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t}^{b-}-\hat{\mathbf x}_{t}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t+1}^{f})^\top 
		\end{equation*}
		 Compute the smoother gain, the smoothed state and the smoothed covariance matrix
		\begin{equation*}
			\mathbf S_t=\mathbf M_{t +1}\left[ \mathbf P_{t +1}^{b-}\right] ^{-1} 
		\end{equation*}
		\begin{equation*}
			\hat{\mathbf x}_t^b=\hat{\mathbf x}_t^f+\mathbf S_t\left[\hat{\mathbf x}_{t+1}^{b}-\hat{\mathbf x}_{t+1}^{b-}\right] 
		\end{equation*}
		\begin{equation*}
			\mathbf P_{t}^{b}=\mathbf P_{t}^{f}+\mathbf S_t\left[\mathbf P_{t+1}^{b}-\mathbf P_{t+1}^{b-} \right]\mathbf S_t^\top 
		\end{equation*}
	\end{algorithmic}
\end{small}
\end{algorithm}

Although the system is nonlinear, the parameters of the system are linear with respect to the state. This is exploited by our procedure where the parameter estimation uses a Least Squares (LS) method that minimises the sum of the squared errors (of a predicted state update) with each new state estimate (see~\ref{LeastSquaresAppendix}).

>>>>

	\label{DiscreteTimeModel} v_{t+T_s}\left(\mathbf{r}\right) &=& \xi v_t\left(\mathbf{r}\right) + T_s \int_\Omega { w\left(\mathbf{r},\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'} \nonumber\\ 

@@

	 w\left(\mathbf{r},\mathbf{r}'\right) =\boldsymbol{\psi}^\top\left(\mathbf{r},\mathbf{r}'\right) \boldsymbol{\theta},

@@

where $\boldsymbol{\psi}(r,r')$ is a vector of Gaussian basis functions and $\boldsymbol{\theta}$ is a vector of scaling parameters. By assuming a Gaussian isotropic connectivity structure the kernel basis functions can be written as $\psi(\mathbf{r}-\mathbf{r}')$. We will assume that we know the parametric form of the connectivity basis functions, where the scaling parameters $\boldsymbol{\theta}$ are unknown. Each connectivity basis function can, individually, be considered a layer in the Wilson and Cowan model, representing short range excitation, surround inhibition and mid-range excitation. Making substitutions of equations~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} into~\ref{DiscreteTimeModel} we get 

@@

Let the spectral representation of the dynamic field at time $t$ be denoted by $V_t(\boldsymbol{\nu})$. Based on Shannon's sampling theorem, for an accurate representation of this field by spatially discrete observations, $V_t(\boldsymbol{\nu})$ needs to be spatially band-limited. Such a condition ensures that $V_t(\boldsymbol{\nu}) \approx 0 ~ \forall \boldsymbol{\nu} > \boldsymbol{\nu}_c$, where $\boldsymbol{\nu}_c$ is a cutoff frequency (typically taken as -3~dB point) with $\boldsymbol{\nu}_c = [\nu_c ~ \nu_c]^\top$. Given such a band-limited field, the distance between adjacent sensors, $\Delta_y$, needs to satisfy 
\begin{equation}
	\label{eq:MinimumSensorDistance} \Delta_y < \frac{1}{2\rho_y\nu_{c}}, 
\end{equation}
where $\rho_y \in \mathbb{R} \ge 1$ is an oversampling parameter. This condition must be satisfied to avoid spectral aliasing effects in the representation of the hidden dynamic field $v_t(\mathbf{r})$ by the sampled observations, $\mathbf{y}_t$.

@@

% 
% Due to the high dimensionality of the brain and the current electrode systems, we can not expect to have more sensors than basis functions.
\section{State and Parameter Estimation}\label{StateAndParameterEstimationSection} In this section we describe the procedure for estimating the states, $\mathbf{x}_t$, and the connectivity kernel parameters, $\boldsymbol \theta$, and the synaptic time constant, $\xi$. The estimation process is a two part iterative algorithm, consisting of a state estimation step followed by a parameter estimation step. At each iteration, the sequence of estimated state vectors is used to update the parameter set. The resulting parameters are then used to estimate a new state vector sequence for the next iteration. The procedure stops when the parameters converge. The algorithm is initialised using a bounded random state vector sequence that guarantees the initial estimated parameter set forms a stable kernel.

The unscented Rauch Tung Striebel smoother (URTSS)~\cite{Sarkka2010} is used for the state estimation. The URTSS incorporates an unscented Kalman Filter (UKF)~\cite{Julier1997, Merwe2003} in a forward iteration to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward iteration to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and the second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. The sigma points, $\mathcal X_i$, are calculated using the unscented transform as follows:
\begin{equation}\label{eq:sigmapoints1}
	\mathcal X_{0}=\bar x 
\end{equation}
\begin{equation}
	\mathcal X_{i}=\bar x+(\sqrt{( L + \lambda)\mathbf P_x})_i, \quad i=1, \dots, L 
\end{equation}
\begin{equation}\label{eq:sigmapoints2}
	\mathcal X_{i}=\bar x-(\sqrt{( L + \lambda)\mathbf P_x})_{i- L}, \quad i= L+1, \dots, 2 L 
\end{equation}
where $\bar x$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$, $\mathbf{P}_x$ is the corresponding covariance matrix from the filtering or smoothing, $(\sqrt{( L + \lambda)\mathbf P_x})_i$ is the $i$th column of the weighted matrix square root of $\mathbf P_x$ and $L$ is the dimension of the state space. The total number of sigma points is $2L+1$. The scaling parameter, $\lambda$, is defined as 
\begin{equation}\label{eq:sigmapoints3}
	\lambda=\alpha^2( L+\kappa) - L 
\end{equation}
where the constant $\alpha$ determines the spread of the sigma points, $\beta$ incorporates prior knowledge of the distribution of $\mathbf{x}$ and $\kappa$ is a scaling parameter (see ~\cite{Haykin2001} for more details).

The sigma vectors are propagated through the system equations and weighted to form the predicted mean and covariance. The weights are calculated by 
	\mathbf W_0^{(m)}=\frac{\lambda}{ L+\lambda} 
	\mathbf W_0^{(c)}=\frac{\lambda}{ L+\lambda}+(1-\alpha^2+\beta) 
	\mathbf W_i^{(m)}=\mathbf W_i^{(c)}=\frac{1}{2( L+\lambda)} \quad i=1, \dots, 2L. 
Since the observation equation is linear (equation~\ref{ObservationEquation}) the standard Kalman Filter update equations are  used to correct the predicted states. The state estimates from the forward filtering are used to form a new set of sigma points for the smoother, as described above. To compute the smoother gain, the predicted cross-covariance matrix of the states is used with the covariance matrix. A summary of the URTSS procedure is given in Algorithm~\ref{UKFAlgorithm}. It should be noted that  the disturbance, $e_t\left(\mathbf{r}\right)$ and the measurement noise, $\boldsymbol{\varepsilon}_t$, are additive. Therefore, the additive form of the URTS is used, otherwise the state vector should be augmented with the noise terms. The additive disturbance form reduces the dimension and the number of sigma points used in the algorithm.
\begin{algorithm}
	\caption{The Unscented RTS Smoother}\label{UKFAlgorithm} 
	\begin{algorithmic}[1] 
		\State Forward initialisation 
		 \hat{\mathbf x}_0, \mathbf P_0 
		\State Forward iteration: for $t \in \left\lbrace 0,\cdots, T\right\rbrace $,
		calculate the sigma points $\mathcal X_{i,t}^f$ using equations \ref{eq:sigmapoints1}-\ref{eq:sigmapoints3} and propagate through equation~\ref{eq:QmatrixForSigmapoints}
% 		\begin{small}
			\mathcal X_{i,t+1}^{f-}=Q(\mathcal X_{i,t}^f) 
% 		\end{small}
		calculate the predicted state and the predicted covariance matrix
			\hat{\mathbf x}_{t+1}^{f-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{f-} 
			\mathbf P_{t +1}^{f-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})^\top+\boldsymbol \Sigma_e 
		compute the filter gain, the filtered state and the filtered covariance matrix using the standard Kalman Filter update equations
			\mathcal K_{t+1}=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_v)^{-1} 
			\hat{\mathbf x}_{t+1}^{f}=\hat{\mathbf x}_{t+1}^{f-}+\mathcal K_{t+1}(\mathbf y_{t+1}-\mathbf C\hat{\mathbf x}_{t +1}^{f-}) 
			\mathbf P_{t+1}^f=(\mathbf I - \mathcal K_{t+1}\mathbf C)\mathbf P_{t +1}^{f-} 
		\State Backward initialisation 
			\mathbf P_T^b= \mathbf P_T^f, \quad \hat{\mathbf x}^b_T= \hat{\mathbf x}^f_T 
		\State Backward iteration: for $t \in \left\lbrace T-1, \cdots, 0 \right\rbrace $ calculate the sigma points $\mathcal X_{i,t}^b$ and propogate them through equation \ref{eq:QmatrixForSigmapoints}
			\mathcal X_{i,t+1}^{b-}=Q(\mathcal X_{i,t}^b,u_t) 
		 compute the predicted state, the predicted covariance matrix and the cross-covariance matrix
			\hat{\mathbf x}_{t+1}^{b-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{b-} 
			\mathbf P_{t +1}^{b-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{f})^\top+\boldsymbol \Sigma_e 
			\mathbf M_{t +1}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t}^{b-}-\hat{\mathbf x}_{t}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t+1}^{f})^\top 
		 Compute the smoother gain, the smoothed state and the smoothed covariance matrix
			\mathbf S_t=\mathbf M_{t +1}\left[ \mathbf P_{t +1}^{b-}\right] ^{-1} 
			\hat{\mathbf x}_t^b=\hat{\mathbf x}_t^f+\mathbf S_t\left[\hat{\mathbf x}_{t+1}^{b}-\hat{\mathbf x}_{t+1}^{b-}\right] 
			\mathbf P_{t}^{b}=\mathbf P_{t}^{f}+\mathbf S_t\left[\mathbf P_{t+1}^{b}-\mathbf P_{t+1}^{b-} \right]\mathbf S_t^\top 
	\end{algorithmic}
\end{algorithm}

Although the system is nonlinear, the parameters of the system are linear with respect to the state. This is exploited by our procedure where the parameter estimation uses a least squares (LS) method that minimises the sum of the squared errors (of a predicted state update) with each new state vector sequence estimate (see~\ref{LeastSquaresAppendix}).
